---
title: "Explore differential methylation"
author: "Alice Balard"
date: '2022-06-23'
output: 
  html_document:
    number_sections: true
    toc: true
    theme: united
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup}
library(knitr)
knitr::opts_chunk$set(cache = TRUE, warning = FALSE,
                      message = FALSE, cache.lazy = FALSE, cache.path = "../../gitignore/RmdCache/", # keep heavy data in gitignore cache
                      fig.path = "Rmdfig/")
```

```{r load, message=FALSE, results='hide'}
machine="mythinkpad" # define the machine we work on
loadALL = TRUE # load all uniteCov objects
loadannot = TRUE # load genome annotations
sourceDMS = TRUE # Load the calculated DMS
sourceSubUnite = FALSE
source("R02.3_DATALOAD.R")

## So far, we don't use DMR (because RRBS single end, this is not super meaningful)
rm(DMRBPlist, DMRlist)
```

**Data getting loaded:**

-   uniteCovALL_woSexAndUnknowChr 55530 CpG positions shared by all
fish
-   uniteCov6_G1_woSexAndUnknowChrOVERLAP 1001880 CpG positions
shared by half the parents in each trt group, overlapping with the
parental ones
-   uniteCov14_G2_woSexAndUnknowChrOVERLAP 1001880 CpG positions
shared by half the offspring in each trt group, overlapping with the
offspring ones

# Global methylation and fitness

Compare fitness traits between the different offsprings groups. Follow up of Sagonas 2020 & Ferre Ortega's master's dissertation

## Calculate BCI

Kaufmann et al. 2014: Body condition of the G2 fish, an estimate of fish health and a predictor of energy reserves and reproductive success, was calculated using there residuals from the regression of body mass on body length (Chellappaet al.1995).

```{r}
fullMetadata_OFFS$BCI <- residuals(lmer(Wnettofin ~ Slfin * Sex + (1|brotherPairID), data=fullMetadata_OFFS))

## and for parents (no sex difference, only males):
fullMetadata_PAR$BCI <- residuals(lmer(Wnettofin ~ Slfin + (1|brotherPairID), data=fullMetadata_PAR))
```

Effect of paternal treatment on body condition of offspring: Kaufmann et al. 2014: "To investigate in which way paternal G1 exposure affected
offspring tolerance, we tested how the relationship between G2 body condition and infection intensity was affected by paternal G1 exposure. This was tested in a linear mixed model on G2 body condition with paternal G1 treatment and the interaction between paternal G1 treatment and G2 infection intensity as fixed effects. Maternal half-sibship identity was set as a random effect"

## Effect of paternal exposure on tolerance

### BCI per trt

Effect of treatment groups of offspring on body condition(Kaufmann et al. 2014): "The linear mixed effect model (nlme function in R) included
G2 body condition as dependent variable, sex, G2 treatment (exposed vs. control), paternal G1 treatment (exposed vs. control) and their interactions as fixed effects as well as maternal G2 half-sibship identity as a random effect"

```{r, fig.width=6, fig.height=6}
mod1 <- lme(BCI ~ offsTrt * patTrt, random=~1|brotherPairID,data=fullMetadata_OFFS)
anova(mod1) # strong significant effect of both offspring trt & paternal + interactions

mod1.2 <- lme(BCI ~  trtG1G2, random=~1|brotherPairID,data=fullMetadata_OFFS)
## pairwise posthoc test
emmeans(mod1.2, list(pairwise ~ trtG1G2), adjust = "tukey")
## Control father - treatment offspring has a strongly significantly lower BC than
## every other group, same as Kaufmann et al. 2014

myplot1 <- ggplot(fullMetadata_OFFS, aes(x=trtG1G2, y = BCI, fill=trtG1G2))+
  geom_boxplot()+
  geom_signif(comparisons = list(c("NE_control", "NE_exposed")),
              map_signif_level=TRUE, annotations="***",
              y_position = 150, tip_length = 0, vjust=0.4) +
  geom_signif(comparisons = list(c("NE_exposed", "E_control")),
              map_signif_level=TRUE, annotations="***",
              y_position = 200, tip_length = 0, vjust=0.4) +
  geom_signif(comparisons = list(c("NE_exposed", "E_exposed")),
              map_signif_level=TRUE, annotations="***",
              y_position = 250, tip_length = 0, vjust=0.4) +
  scale_fill_manual(values = colOffs)+
  theme_bw() + theme(legend.position = "none") +
  ylab("Body Condition Index") +
  scale_x_discrete(labels=c("NE_control" = "G1 control\nG2 control", "NE_exposed" = "G1 control\nG2 infected",
                            "E_control" = "G1 infected\nG2 control", "E_exposed" = "G1 infected\nG2 infected"),
                   name = NULL)
myplot1
```

### Tolerance (slope of BCI on worms count) per paternal status

```{r, fig.width=6,fig.height=4}
mod_Tol <- lmer(BCI ~ No.Worms*PAT + (1|brotherPairID)+ (1|Sex), data=fullMetadata_OFFS, REML = F)

## Model selection:
step(mod_Tol, reduce.random = F) # Model found: full model

## The slope of BCI on nbrworms varies upon treatment
plot(ggpredict(mod_Tol, terms = c("No.Worms", "PAT")), add.data=T)+ theme_bw() +
  ylab("Body Condition Index") + xlab("Number of worms") +
  ggtitle("Predicted values of Body Condition Index in offspring")+
  scale_color_manual(NULL, values = c("black", "red")) +
  scale_fill_manual(NULL, values = c("black", "red"))  +
  scale_x_continuous(breaks = 0:10)+
  geom_point(size=0)+ # to have color key in legend as point
  guides(colour = guide_legend(override.aes = list(size=3,linetype=0, fill = NA)))
```

## Link between methylation and fitness (BCI and tolerance)

Calculate number of methylated sites, mean coverage, and residuals of methylated sites by covered sites (to account for coverage bias)

```{r}
mycalcRMS <- function(myUniteCov, myMetaData){
  percMethMat = methylKit::percMethylation(myUniteCov)
  # create a dataframe with all info
  percMethDF = data.frame(SampleID = colnames(percMethMat),
                          Nbr_methCpG = colSums(percMethMat>=70 & !is.na(percMethMat)), ## number of methylated sites
                          Nbr_coveredCpG = colSums(!is.na(percMethMat)), ## number of sites covered in this sample
                          Nbr_NOTcoveredCpG = colSums(is.na(percMethMat)),## number of sites NOT covered in this sample
                          MeanCoverage = colMeans(methylKit::getData(myUniteCov)[,myUniteCov@coverage.index], na.rm = T), ## coverage.index: vector denoting which columns in the data correspond to coverage values
                          OverallPercentageMethylation = colMeans(methylKit::percMethylation(myUniteCov), na.rm = T))
  
  ## RMS in this sample based on covered sites
  percMethDF$RMS_coveredCpG = percMethDF$Nbr_methCpG / percMethDF$Nbr_coveredCpG
  ## merge with original metadata:
  myMetaData = merge(myMetaData, percMethDF)
  # calculate also RMS global, considering CpG covered or not (to compare)
  myMetaData$RMS_allCpG_coveredOrNot = myMetaData$Nbr_methCpG / (myMetaData$M.Seqs_rawReads*10e6)
  # calculate residuals of nbr of methCpG by nbr of covered CpG
  myMetaData$res_Nbr_methCpG_Nbr_coveredCpG = residuals(
    lm(myMetaData$Nbr_methCpG ~ myMetaData$Nbr_coveredCpG))
  ## REORDER myMetaData by sample ID
  myMetaData = myMetaData[order(as.numeric(gsub("S", "", myMetaData$SampleID))),]
  return(myMetaData)
}

fullMetadata <- mycalcRMS(uniteCovALL_woSexAndUnknowChr, fullMetadata)

fullMetadata_PAR <- mycalcRMS(uniteCov6_G1_woSexAndUnknowChrOVERLAP, fullMetadata_PAR)

fullMetadata_OFFS  <- mycalcRMS(uniteCov14_G2_woSexAndUnknowChrOVERLAP, fullMetadata_OFFS)
```

## Nbr/Ratio of Methylated Sites in different groups

We kept in total `r nrow(fullMetadata)` samples. On average, `r round(mean(fullMetadata$M.Seqs_rawReads), 2)` [+/- `r round(qnorm(0.975)*sd(fullMetadata$M.Seqs_rawReads)/sqrt(nrow(fullMetadata)), 2)`] million reads were sequenced. The average mapping efficiency was `r round(mean(fullMetadata$MappingEfficiency.BSBoldvsGynogen))`% [+/- `r round(qnorm(0.975)*sd(fullMetadata$MappingEfficiency.BSBoldvsGynogen)/sqrt(nrow(fullMetadata)),2)`%].

The mean coverage per CpG site in the full dataset, considering positions covered in all fish, is `r round(mean(fullMetadata$MeanCoverage))` [+/- `r round(qnorm(0.975)*sd(fullMetadata$MeanCoverage)/sqrt(nrow(fullMetadata)),2)`].

**For G1 only:** We kept in total `r nrow(fullMetadata_PAR)` samples. On average, `r round(mean(fullMetadata_PAR$M.Seqs_rawReads), 2)` [+/- `r round(qnorm(0.975)*sd(fullMetadata_PAR$M.Seqs_rawReads)/sqrt(nrow(fullMetadata_PAR)), 2)`] million reads were sequenced. The average mapping efficiency was
# `r round(mean(fullMetadata_PAR$MappingEfficiency.BSBoldvsGynogen))`% [+/- `r round(qnorm(0.975)*sd(fullMetadata_PAR$MappingEfficiency.BSBoldvsGynogen)/sqrt(nrow(fullMetadata_PAR)),2)`%].

The mean coverage per CpG site in G1, considering positions shared by at least 6 fish per treatment group (half individuals per group), and which overlap with positions retained in G2, is `r round(mean(fullMetadata_PAR$MeanCoverage))` [+/- `r round(qnorm(0.975)*sd(fullMetadata_PAR$MeanCoverage)/sqrt(nrow(fullMetadata_PAR)),2)`].

**For G2 only:** We kept in total `r nrow(fullMetadata_OFFS)` samples. On average, `r round(mean(fullMetadata_OFFS$M.Seqs_rawReads), 2)` [+/- `r round(qnorm(0.975)*sd(fullMetadata_OFFS$M.Seqs_rawReads)/sqrt(nrow(fullMetadata_OFFS)), 2)`] million reads were sequenced. The average mapping efficiency was
`r round(mean(fullMetadata_OFFS$MappingEfficiency.BSBoldvsGynogen))`% [+/- `r round(qnorm(0.975)*sd(fullMetadata_OFFS$MappingEfficiency.BSBoldvsGynogen)/sqrt(nrow(fullMetadata_OFFS)),2)`%].

The mean coverage per CpG site in G2, considering positions shared by at least 14 fish per treatment group (half individuals per group), and which overlap with positions retained in G1, is `r round(mean(fullMetadata_OFFS$MeanCoverage))` [+/- `r round(qnorm(0.975)*sd(fullMetadata_OFFS$MeanCoverage)/sqrt(nrow(fullMetadata_OFFS)),2)`].

## Choice of global methylation value

```{r}
cor.test(fullMetadata_PAR$Nbr_coveredCpG,
         fullMetadata_PAR$Nbr_methCpG, method = "spearman")
## S = 350, p-value = 2.15e-06, rho = 0.85
ggplot(fullMetadata_PAR, aes(x=Nbr_coveredCpG, y=Nbr_methCpG))+
  geom_smooth(method = "lm", col="black")+
  geom_point(aes(col=trtG1G2), size = 3)+ scale_color_manual(values = c("grey", "red")) +
  theme_bw() + ggtitle(label = "Parents, CpG shared by half fish/trt")

## Check after RMS correction for coverage bias: CORRECTED (p-value = 0.4485)
cor.test(fullMetadata_PAR$Nbr_coveredCpG,
         fullMetadata_PAR$RMS_coveredCpG, method = "spearman")
ggplot(fullMetadata_PAR, aes(x=Nbr_coveredCpG, y=RMS_coveredCpG))+
  geom_smooth(method = "lm", col="black")+
  geom_point(aes(col=trtG1G2), size = 3)+ scale_color_manual(values = c("grey", "red")) +
  theme_bw() + ggtitle(label = "Parents, CpG shared by half fish/trt")

## and with residuals: COMPLETELY CORRECTED p-value = 0.9562
cor.test(fullMetadata_PAR$Nbr_coveredCpG,
         fullMetadata_PAR$res_Nbr_methCpG_Nbr_coveredCpG, method = "spearman")
ggplot(fullMetadata_PAR, aes(x=Nbr_coveredCpG, y=res_Nbr_methCpG_Nbr_coveredCpG))+
  geom_smooth(method = "lm", col="black")+
  geom_point(aes(col=trtG1G2), size = 3)+ scale_color_manual(values = c("grey", "red")) +
  theme_bw() + ggtitle(label = "Parents, CpG shared by half fish/trt")

############
## Offspring:
cor.test(fullMetadata_OFFS$Nbr_coveredCpG,
         fullMetadata_OFFS$Nbr_methCpG, method = "spearman")
## S = 20254, p-value < 2.2e-16 rho = 0.91
ggplot(fullMetadata_OFFS, aes(x=Nbr_coveredCpG, y=Nbr_methCpG))+
  geom_smooth(method = "lm", col="black")+
  geom_point(aes(col=trtG1G2), size = 3)+ scale_color_manual(values = colOffs) +
  scale_x_continuous("Number of cytosines covered") +
  scale_y_continuous("Number of methylated cytosines") +
  theme_bw() + ggtitle(label = "Offspring, CpG shared by half fish/trt")

## Plot distance to residuals:
fit <- lm(Nbr_methCpG ~ Nbr_coveredCpG, data = fullMetadata_OFFS)
plotdf <- fullMetadata_OFFS
plotdf$predicted <- predict(fit)   # Save the predicted values
plotdf$residuals <- residuals(fit)
ggplot(plotdf, aes(x=Nbr_coveredCpG, y=Nbr_methCpG))+
  geom_smooth(method = "lm", col="black")+
  geom_segment(aes(xend = Nbr_coveredCpG, yend = predicted), col = "grey") +
  geom_point(aes(col=trtG1G2), size = 3)+ scale_color_manual(values = colOffs) +
  scale_x_continuous("Number of cytosines covered") +
  scale_y_continuous("Number of methylated cytosines") +
  theme_bw() + ggtitle(label = "Offspring, CpG shared by half fish/trt")

## Check after RMS correction for coverage bias: SEMI CORRECTED (p-value = 0.01, rho = -0.24)
cor.test(fullMetadata_OFFS$Nbr_coveredCpG,
         fullMetadata_OFFS$RMS_coveredCpG, method = "spearman")
ggplot(fullMetadata_OFFS, aes(x=Nbr_coveredCpG, y=RMS_coveredCpG))+
  geom_point(aes(col=trtG1G2), size = 3)+ scale_color_manual(values = colOffs) +
  geom_smooth(method = "lm", col="black")+
  theme_bw() + ggtitle(label = "Offspring, CpG shared by half fish/trt")

## and with residuals: COMPLETELY CORRECTED p-value = 0.51
cor.test(fullMetadata_OFFS$Nbr_coveredCpG,
         fullMetadata_OFFS$res_Nbr_methCpG_Nbr_coveredCpG, method = "spearman")
ggplot(fullMetadata_OFFS, aes(x=Nbr_coveredCpG, y=res_Nbr_methCpG_Nbr_coveredCpG))+
  geom_point(aes(col=trtG1G2), size = 3)+ scale_color_manual(values = colOffs) +
  geom_smooth(method = "lm", col="black")+
  scale_x_continuous("Number of cytosines covered") +
  scale_y_continuous("Residuals of number of methylated cytosines\n on number of cytosines covered") +
  theme_bw() + ggtitle(label = "Offspring, CpG shared by half fish/trt")
```

## Why we should we correct for sex

### No difference in mappability (p\>0.05)

```{r}
mod = lm(MappingEfficiency.BSBoldvsGynogen ~ Sex, data = fullMetadata_OFFS)
summary(step(mod))
plot(ggpredict(mod, terms = c("Sex")), add.data = T)
```

NB: this is WITH unknown and sex chromosomes, before filtering.

### No difference in number of reads (p\>0.05)

```{r}
mod = lm(M.Seqs_rawReads ~ Sex, data = fullMetadata_OFFS)
summary(step(mod))
plot(ggpredict(mod, terms = c("Sex")), add.data = T)
```
NB: this is WITH unknown and sex chromosomes, before filtering.

### No difference in mean coverage per CpG in the filtered dataset (p\>0.05)

```{r}
mod = lm(MeanCoverage ~ Sex, data = fullMetadata_OFFS)
summary(step(mod))
plot(ggpredict(mod, terms = c("Sex")), add.data = T)
```

NB: this is in G2, considering positions shared by at least 14 fish per treatment group (half individuals per group), and which overlap with positions retained in G1, without sex and unknown chromosome (after filtering) 

### No difference in number of sites covered in the filtered dataset (p\>0.05)

```{r}
mod = lm(Nbr_coveredCpG ~ Sex, data = fullMetadata_OFFS)
summary(step(mod))
plot(ggpredict(mod, terms = c("Sex")), add.data = T)
```

NB: this is in G2, considering positions shared by at least 14 fish per treatment group (half individuals per group), and which overlap with positions retained in G1, without sex and unknown chromosome (after filtering)

### No difference in number of sites covered in the filtered dataset (p\>0.05)

```{r}
mod = lm(OverallPercentageMethylation ~ Sex, data = fullMetadata_OFFS)
summary(step(mod))
plot(ggpredict(mod, terms = c("Sex")), add.data = T)
```

NB: this is in G2, considering positions shared by at least 14 fish per treatment group (half individuals per group), and which overlap with
positions retained in G1, without sex and unknown chromosome (after filtering)

### Males have a lower global methylation than females (residuals of nbr of methylated sites by nbr of sites covered)

```{r, fig.width=5, fig.height=4}
mod = lm(res_Nbr_methCpG_Nbr_coveredCpG ~ Sex, data = fullMetadata_OFFS)
summary(step(mod)) # sex is significant p = 0.000157 ***
anova(mod)

plot(ggpredict(mod, terms = c("Sex")), add.data = T) +
  xlab(NULL)+
  ylab("Residuals of N methylated sites on N covered sites") +
  ggtitle("Predicted values of global methylation in offspring")
```

## Are mean residuals meth sites different following tolerance slope?

### In all offspring

```{r}
fullMetadata_OFFS$res_Nbr_methCpG_Nbr_coveredCpG_div1000 <- (fullMetadata_OFFS$res_Nbr_methCpG_Nbr_coveredCpG)/1000

mod_Tol.Meth <- lmer(BCI ~ res_Nbr_methCpG_Nbr_coveredCpG_div1000*No.Worms*PAT + (1|brotherPairID)+ (1|Sex),
                     data=fullMetadata_OFFS, REML = F)

## Model selection:
step(mod_Tol.Meth, reduce.random = F) # Model found: BCI ~ No.Worms + PAT + (1 | brotherPairID) + (1 | Sex) + No.Worms:PAT
## The slope of BCI on nbrworms varies upon treatment but methylation does NOT vary with tolerance
mod_Tol.Meth <- lmer(BCI ~ No.Worms*PAT + (1|brotherPairID)+ (1|Sex),
                     data=fullMetadata_OFFS)

## And by treatment instead of No.worms?
mod_Tol.Meth2 <- lmer(BCI ~ res_Nbr_methCpG_Nbr_coveredCpG_div1000*PAT*outcome + (1|brotherPairID)+ (1|Sex),
                      data=fullMetadata_OFFS, REML = F)

## Model selection:
step(mod_Tol.Meth2, reduce.random = F) # Model found: BCI ~ PAT + outcome + (1 | brotherPairID) + (1 | Sex) + PAT:outcome
```

The slope of BCI on nbr worms varies upon parental treatment, but methylation does NOT vary with tolerance

### In exposed offspring only

```{r}
## By group, tolerance slope as a function of methylation residuals:
modFULL <- lmer(BCI ~ res_Nbr_methCpG_Nbr_coveredCpG_div1000*No.Worms + (1|brotherPairID) + (1|Sex),
                data = fullMetadata_OFFS[fullMetadata_OFFS$trtG1G2 %in% c("NE_exposed", "E_exposed"),])
## Model selection:
step(modFULL, reduce.random = F) # Model found: BCI ~ (1 | brotherPairID) + (1 | Sex)

modFULL <- lmer(BCI ~ res_Nbr_methCpG_Nbr_coveredCpG_div1000*PAT + (1|brotherPairID) + (1|Sex),
                data = fullMetadata_OFFS[fullMetadata_OFFS$trtG1G2 %in% c("NE_exposed", "E_exposed"),])
## Model selection:
step(modFULL, reduce.random = F) # Model found: BCI ~ PAT + (1 | brotherPairID) + (1 | Sex)
```

## PCA based on all methylation values in G2

```{r, warning=FALSE}
# 1. get raw values
percmeth = percMethylation(uniteCov14_G2_woSexAndUnknowChrOVERLAP)

# Run PCA on complete data (CpG covered in all fish)
PCA_allpos <- myPCA(x = t(na.omit(percmeth)), incomplete = FALSE)
```

We perform a PCA on `r nrow(na.omit(percmeth))` CpG sites covered in all
G2 individuals. We first perform a test on the complete dataset.

### PCA plot with associated colors for treatments

```{r}
fviz_pca_ind(PCA_allpos$res.PCA,  label="none", habillage=PCA_allpos$metadata$trtG1G2) +
  scale_color_manual(values = colOffs)+
  scale_shape_manual(values=c(19,19,19,19))
```

### PCA plot with associated colors for brother pair

```{r}
fviz_pca_ind(PCA_allpos$res.PCA,  label="none", habillage=as.factor(PCA_allpos$metadata$brotherPairID))
```

```{r}
# The function dimdesc() can be used to identify the most correlated variables with a given principal component.
mydimdesc <- dimdesc(PCA_allpos$res.PCA, axes = c(1,2), proba = 0.05)
```

There are `r nrow(mydimdesc$Dim.1$quanti)` CpG sites most correlated (p
\< 0.05) with the first principal component , and
`r nrow(mydimdesc$Dim.2$quanti)` with the second principal component.

The 2 first PCA axes do not explain BCI (p\<0.05)

### How much of the BCI variance is explained by each variables?

```{r}
# Percentage of variance explained by each factor:
formula(PCA_allpos$modSel) # BCI ~ No.Worms + PAT + (1 | brotherPairID) + (1 | Sex) + No.Worms:PAT
mod_noworms = lmer(BCI ~ PAT + (1 | brotherPairID) + (1 | Sex), data = PCA_allpos$metadata)
mod_noPAT = lmer(BCI ~ No.Worms + (1 | brotherPairID) + (1 | Sex), data = PCA_allpos$metadata)

# R2c conditional R2 value associated with fixed effects plus the random effects.
A = (MuMIn::r.squaredGLMM(PCA_allpos$modSel)[2] -
       MuMIn::r.squaredGLMM(mod_noworms)[2])*100

B = (MuMIn::r.squaredGLMM(PCA_allpos$modSel)[2] -
       MuMIn::r.squaredGLMM(mod_noPAT)[2])*100
```

-   `r round(A, 2)`% of the variance in associated with the parasite
load (number of worms)
-   `r round(B, 2)`% of the variance in associated with the paternal
treatment

## Pretty summary picture

```{r}
# Set up scatterplot
scatterplot <- ggplot(fullMetadata_OFFS,
                      aes(x = res_Nbr_methCpG_Nbr_coveredCpG,
                          y = BCI, fill=trtG1G2)) +
  geom_point(pch=21, size =3, alpha = .8) +
  guides(color = "none") +
  scale_fill_manual(values = colOffs, name = "Treatment",
                    labels = c("G1 control - G2 control", "G1 control - G2 exposed", "G1 exposed - G2 control", "G1 exposed - G2 exposed")) +
  theme(plot.margin = margin()) + theme_bw() +
  theme(legend.position = "none") +
  xlab("Methylation residuals (methylated sites/coverage")+
  ylab("Body Condition Index")

# Define marginal histogram
marginal_distribution <- function(x, var, group) {
  ggplot(x, aes_string(x = var, fill = group)) +
    # geom_histogram(bins = 30, alpha = 0.4, position = "identity") +
    geom_density(alpha = 0.6, size = 0.2) +
    guides(fill = "none") +
    scale_fill_manual(values = colOffs) +
    theme_void() +
    theme(plot.margin = margin())
}

# Set up marginal histograms
x_hist <- marginal_distribution(fullMetadata_OFFS, "res_Nbr_methCpG_Nbr_coveredCpG", "trtG1G2")
y_hist <- marginal_distribution(fullMetadata_OFFS, "BCI", "trtG1G2") +
  coord_flip()

# Align histograms with scatterplot
aligned_x_hist <- align_plots(x_hist, scatterplot, align = "v")[[1]]
aligned_y_hist <- align_plots(y_hist, scatterplot, align = "h")[[1]]

# Arrange plots
cowplot::plot_grid(
  aligned_x_hist, NULL, scatterplot, aligned_y_hist, ncol = 2, nrow = 2, rel_heights = c(0.2, 1), rel_widths = c(1, 0.2)
)
```

# Methylation profile

Methylation profiles, CpG present in all fish

## Dendogram of methylations

Methylation profile for the `r nrow(uniteCovALL_woSexAndUnknowChr)` CpG
sites covered in all G1 & G2 fish (N =
`r length(uniteCovALL_woSexAndUnknowChr@sample.ids)`:

```{r, fig.width=15, fig.height=8}
makePrettyMethCluster(uniteCovALL_woSexAndUnknowChr, fullMetadata,
                      my.cols.trt=c("#333333ff","#ff0000ff","#ffe680ff","#ff6600ff","#aaccffff","#aa00d4ff"),
                      my.cols.fam = c(1:4), nbrk = 8)
```

### Offspring:

Methylation profile for the `r nrow(uniteCovALL_G2_woSexAndUnknowChr)`
CpG sites covered in all G2 fish (N =
`r length(uniteCovALL_G2_woSexAndUnknowChr@sample.ids)`:

```{r, fig.width=10, fig.height=4}
makePrettyMethCluster(uniteCovALL_G2_woSexAndUnknowChr, fullMetadata_OFFS,
                      my.cols.trt=c("#ffe680ff","#ff6600ff", "#aaccffff", "#aa00d4ff"),
                      my.cols.fam = c(1:4), nbrk = 8)

# Save
pdf(file = "../../dataOut/clusterALLCpG_offspings.pdf", width = 10, height = 4)
makePrettyMethCluster(uniteCovALL_G2_woSexAndUnknowChr, fullMetadata_OFFS,
                      my.cols.trt=c("#ffe680ff","#ff6600ff", "#aaccffff", "#aa00d4ff"),
                      my.cols.fam = c(1:4), nbrk = 8)
dev.off()
```

## Adonis tests: impact of different variables on methylation pattern

Permutational multivariate analysis of variance (PERMANOVA) is a
non-parametric multivariate statistical test. It is used to compare
groups of objects and test the null hypothesis that the centroids and
dispersion of the groups as defined by measure space are equivalent for
all groups.

```{r}
# make distance matrix with B-C distances
data.dist = makeDatadistFUN(uniteCovALL_G2_woSexAndUnknowChr)

## Adonis test: importance of each predictor
adonis2(data.dist ~ PAT * outcome * Sex * brotherPairID, data = fullMetadata_OFFS)
```

Results: family of the father (brotherPairID) explains more than 14% of
the variance in methylation.

To focus on G1 and G2 treatments, we define the permutation structure
considering brother pairs (N = 8), and use a PERMANOVA to test the
hypothesis that paternal treatment, offspring treatment and their
interactions significantly influencing global methylation.

```{r}
perm <- how(nperm = 1000) # 1000 permutations
setBlocks(perm) <- with(fullMetadata_OFFS, brotherPairID) # define the permutation structure considering brotherPairID and sex
print(adonis2(data.dist ~ PAT * outcome * Sex, data = fullMetadata_OFFS, permutations = perm))
```

Results:

-   1.5% of the variation explained by PAT (R2=0.01470, p \< 0.001)
-   1% of the variation explained by outcome (R2=0.01009, p = 0.03)
-   1.3% of the variation explained by Sex (R2=0.01277, p \< 0.01)

## NMDS

```{r, results='hide'}
#### RUN Goodness of fit
myGOF.NMDS.FUN(dataset = uniteCovALL_G2_woSexAndUnknowChr)
```

Goodness of fit for NMDS suggested the presence of six dimensions with a
stress value \> 0.1 and 2 with \> 0.2

```{r, results='hide'}
## to find the seed that allows convergence:
# sapply(3:10, function(x) myNMDS(dataset = uniteCovALL_G2_woSexAndUnknowChr, metadata = fullMetadata_OFFS, myseed = x))
NMDSanalysis <- myNMDSFUN(dataset = uniteCovALL_G2_woSexAndUnknowChr, metadata = fullMetadata_OFFS, myseed = 4)
```

```{r, fig.height = 11, fig.width = 10}
NMDSanalysis$NMDSplot
# Save
pdf(file = "../../dataOut/NMDSplot_allG2.pdf", width = 9, height = 11)
NMDSanalysis$NMDSplot
dev.off()
```

## The methylation pattern is more affected by direct treatment when the father was infected (Adonis test WITHIN both parental trt)

### 1. Parents NOT infected

```{r}
AdonisWithinG1trtFUN(trtgp = c(2,3))
```

Results:

-   brother pair explains 22% of the observed variance p \< 0.01
-   sex explains 2.3% of the observed variance p \< 0.001
-   Offspring trt explains 1.6% of the observed variance (non
significant)

```{r, results="hide"}
NMDSanalysis_G1control <- myNMDSFUN(dataset = uniteCovALL_G2_woSexAndUnknowChr,
                                    metadata = fullMetadata_OFFS, myseed = 25,
                                    byParentTrt=TRUE,
                                    trtgp = c(2,3))
```

```{r,  fig.height = 9, fig.width = 10}
#png(filename = "../../dataOut/NMDSplot_G1fromControlG2.png", width = 900, height = 900)
NMDSanalysis_G1control$NMDSplot
#dev.off()
```

### 2. Parents infected

```{r}
AdonisWithinG1trtFUN(trtgp = c(5,6))
```

Results:

-   brother pair explains 22% of the observed variance p = 0.01
-   sex explains 2.1% of the observed variance (non significant)
-   Offspring trt explains 2.3% of the observed variance p \< 0.01

```{r, results="hide"}
NMDSanalysis_G1infected <- myNMDSFUN(dataset = uniteCovALL_G2_woSexAndUnknowChr,
                                     metadata = fullMetadata_OFFS, myseed = 25,
                                     byParentTrt=TRUE,
                                     trtgp = c(5,6))
```

```{r,  fig.height = 9, fig.width = 10}
#png(filename = "../../dataOut/NMDSplot_G1fromInfectedG2.png", width = 900, height = 900)
NMDSanalysis_G1infected$NMDSplot
#dev.off()
```

# Differential methylations "Bottom up": DM in offspring

## Description of DMS in the four different comparisons

Differential methylation by brother pair (sex as covariate):

1.  CC-TC = CONTROL fish (parent CvsT)
2.  CT-TT = TREATMENT fish (parent CvsT)
3.  CC-CT = fish from CONTROL parents (G2 CvsT)
4.  TC-TT = fish from TREATMENT parents (G2 CvsT)

### Attribute plot of the DMS in every BP for the 4 comparisons

```{r}
## All DMS are stored in DMSBPlist by brother pair

## Vector of all 4 comparisons
vecCompa <- c("CC_TC", "CT_TT", "CC_CT", "TC_TT")
vecCompaVerbose <- c("Control offspring in control vs infected parents", "Infected offspring in control vs infected parents", "Control vs infected offspring from control parent", "Control vs infected offspring from infected parent") # this is useful in plots

## Extract DMS for all 4 comparisons (by position)
myPosList = lapply(DMSBPlist, lapply, function(x){paste(x$chr, x$end)})

## Prepare df for complexUpset
getUpsetDF = function(i, NBP){ # for a given comparison
  A = get2keep(vecCompa[i], NBP)
  A2 = lapply(A, function(x){
    x = data.frame(x)    # vector of DMS as df
    names(x) = "DMS"    # name each CpG
    return(x)
  })
  ## Add BP name
  for (i in 1:length(names(A2))){
    A2[[i]]["BP"] = names(A2)[i]
  }
  # make a dataframe
  A2 = A2 %>% reduce(full_join, by = "DMS")
  # names column with BP id
  for (i in 2:ncol(A2)) {names(A2)[i] = unique(A2[!is.na(A2[i]), i])}
  # replace by 0 or 1 the DMS absence/presence
  A = data.frame(apply(A2[2:9], 2, function(x) ifelse(is.na(x), 0, 1)))
  # add DMS
  A$DMS = A2$DMS
  return(A)
}
```

Make upset plots for DMS found in at least 6 BP:

```{r, fig.width=10, fig.height=5}
for (i in 1:4){
  df = getUpsetDF(i, NBP = 6)
  print(ComplexUpset::upset(
    df,
    names(df)[1:8],
    width_ratio=0.1,
    themes=upset_default_themes(text=element_text(size=15)),
    sort_intersections_by=c('degree', 'cardinality'),
    queries=query_by_degree(
      df,  names(df)[1:8],
      params_by_degree=list(
        '1'=list(color='red', fill='red'),
        '2'=list(color='purple', fill='purple'),
        '3'=list(color='blue', fill='blue'),
        '4'=list(color='grey', fill='grey'),
        '5'=list(color='red', fill='red'),
        '6'=list(color='purple', fill='purple'),
        '7'=list(color='blue', fill='blue'),
        '8'=list(color='green', fill='green')
      ),
      only_components=c("intersections_matrix", "Intersection size")
    )) + ggtitle(paste0("Differentially methylated sites found in more than six brother pairs in the comparison: \n", vecCompaVerbose[i]))) #+ theme(plot.title = element_text(size = 30)))
}
```

## Statistical setup

PARENTAL effect: DMS found in either CC-TC or CT-TT comparisons
OFFSPRING effect: DMS found in either CC-CT or TC-TT comparisons
INTERACTION effects: DMS found in CC-CT which show a differential methylation (not necessarily significant)
in the opposite direction in TC-TT, or inversely

### Identify the different DMS groups

```{r}
## Subselect those DMS present in at least 4 out of 8 BP
get2keep = function(Compa, NBP = 4){
  x <- lapply(myPosList, function(x){unlist(x[[paste0("DMS_15pc_BP_", Compa)]])})
  f <- table(unlist((x))) # each DMS present between 1 and 8 times
  tokeep <- names(f)[f >= NBP]
  # print(length(tokeep))
  ## Keep the DMS present in 4 families minimum
  DMSBPlist_INTER4 <- lapply(x, function(x){x[x %in% tokeep]})
  ## Reorder by family:
  DMSBPlist_INTER4 <- DMSBPlist_INTER4[names(DMSBPlist_INTER4)[order(names(DMSBPlist_INTER4))]]
  return(DMSBPlist_INTER4)
}

# PARENTAL effect: DMS found in either CC-TC or CT-TT comparisons
DMS_PaternalEffect_4BPmin = unique(c(unlist(get2keep(vecCompa[1], NBP = 4)), unlist(get2keep(vecCompa[2], NBP = 4))))

# OFFSPRING effect: DMS found in either CC-CT or TC-TT comparisons
DMS_OffspringEffect_4BPmin = unique(c(unlist(get2keep(vecCompa[3], NBP = 4)), unlist(get2keep(vecCompa[4], NBP = 4))))

# INTERACTION effects: DMS found in CC-CT which show a differential methylation in the opposite direction in TC-TT, or inversely (reaction norm are inversed)
setDMSInversionSlope = getInteractionDMS() # in customfunctions.R

# Plot a Venn diagram
ggVennDiagram(list("Paternal effect" = DMS_PaternalEffect_4BPmin, "Offspring effect" = DMS_OffspringEffect_4BPmin, "InversionSlope" = setDMSInversionSlope),
              label_alpha = 0) +
  scale_fill_gradient(low="white",high = "red") + theme(legend.position = "none")

# Save:
pdf(file = "../../dataOut/DMS3groupsVenn.pdf", width = 7, height = 6)
ggVennDiagram(list("Paternal effect" = DMS_PaternalEffect_4BPmin, "Offspring effect" = DMS_OffspringEffect_4BPmin, "InversionSlope" = setDMSInversionSlope),
              label_alpha = 0) +
  scale_fill_gradient(low="white",high = "red") + theme(legend.position = "none")
dev.off()

# The five cases of Venn diagram:
caseVennG1only = DMS_PaternalEffect_4BPmin[!DMS_PaternalEffect_4BPmin %in% DMS_OffspringEffect_4BPmin]
caseVennG2only = DMS_OffspringEffect_4BPmin[!DMS_OffspringEffect_4BPmin %in% DMS_PaternalEffect_4BPmin & 
                                              !DMS_OffspringEffect_4BPmin %in% setDMSInversionSlope]
caseVennG1G2NOinter = DMS_OffspringEffect_4BPmin[DMS_OffspringEffect_4BPmin %in% DMS_PaternalEffect_4BPmin & 
                                                   !DMS_OffspringEffect_4BPmin %in% setDMSInversionSlope]
caseVennG1G2inter = DMS_OffspringEffect_4BPmin[DMS_OffspringEffect_4BPmin %in% DMS_PaternalEffect_4BPmin & 
                                                 DMS_OffspringEffect_4BPmin %in% setDMSInversionSlope] 
caseVennG2interNOG1 = DMS_OffspringEffect_4BPmin[!DMS_OffspringEffect_4BPmin %in% DMS_PaternalEffect_4BPmin &
                                                   DMS_OffspringEffect_4BPmin %in% setDMSInversionSlope]
```

The size of the different sections of the Venn diagram are as follows:

- length(caseVennG1only): `r length(caseVennG1only)`
- length(caseVennG2only): `r length(caseVennG2only)`
- length(caseVennG1G2NOinter): `r length(caseVennG1G2NOinter)`
- length(caseVennG1G2inter): `r length(caseVennG1G2inter)`
- length(caseVennG2interNOG1): `r length(caseVennG2interNOG1)`

Plot the reaction norms with real data as a validation of principle:

```{r}
plots = makePlotsobservedReactionNorms()
```

```{r, fig.height=10, fig.width=15}
pdf("../../dataOut/caseVennG1only.pdf", width = 15, height = 10)
plots$plot1
dev.off()
pdf("../../dataOut/caseVennG2only.pdf", width = 15, height = 10)
plots$plot2
dev.off()
pdf("../../dataOut/caseVennG1G2NOinter.pdf", width = 15, height = 10)
plots$plot3
dev.off()
pdf("../../dataOut/caseVennG1G2inter.pdf", width = 15, height = 10)
plots$plot4
dev.off()
pdf("../../dataOut/caseVennG2interNOG1.pdf", width = 15, height = 10)
plots$plot5
dev.off()
```

### Annotate the different DMS groups

```{r annotateEffects}
## Venn diagram by genes found in each comparisons
# PARENTAL effect: DMS found in either CC-TC or CT-TT comparisons
DMS_PaternalEffect_4BPmin_GENES = getAnnotationFun(DMSdf = DMS_PaternalEffect_4BPmin,
                         annotBed12 = annotBed12, annotGff3 = annotGff3, isDMDaDataframeWithBP = FALSE)

# OFFSPRING effect: DMS found in either CC-CT or TC-TT comparisons
DMS_OffspringEffect_4BPmin_GENES = getAnnotationFun(DMSdf = DMS_OffspringEffect_4BPmin,
                         annotBed12 = annotBed12, annotGff3 = annotGff3, isDMDaDataframeWithBP = FALSE)

# INTERACTION effects: DMS found in CC-CT which show a differential methylation in the opposite direction in TC-TT, or inversely (reaction norm are inversed)
setDMSInversionSlope_GENES  = getAnnotationFun(DMSdf = setDMSInversionSlope,
                         annotBed12 = annotBed12, annotGff3 = annotGff3, isDMDaDataframeWithBP = FALSE)

# Plot a Venn diagram
ggVennDiagram(list("Paternal effect" = DMS_PaternalEffect_4BPmin_GENES$Name, "Offspring effect" = DMS_OffspringEffect_4BPmin_GENES$Name,
                   "InversionSlope" = setDMSInversionSlope_GENES$Name),
              label_alpha = 0) +
  scale_fill_gradient(low="white",high = "red") + theme(legend.position = "none") + ggtitle("Genes in each effect")
# Save:
pdf(file = "../../dataOut/DMS3groupsVenn_GENES.pdf", width = 7, height = 6)
ggVennDiagram(list("Paternal effect" = DMS_PaternalEffect_4BPmin_GENES$Name, "Offspring effect" = DMS_OffspringEffect_4BPmin_GENES$Name,
                   "InversionSlope" = setDMSInversionSlope_GENES$Name),
              label_alpha = 0) +
  scale_fill_gradient(low="white",high = "red") + theme(legend.position = "none") + ggtitle("Genes in each effect")
dev.off()

# The five cases of Venn diagram:
annot_caseVennG1only = DMS_PaternalEffect_4BPmin_GENES[!DMS_PaternalEffect_4BPmin_GENES$Name %in% DMS_OffspringEffect_4BPmin_GENES$Name,]
annot_caseVennG2only = DMS_OffspringEffect_4BPmin_GENES[!DMS_OffspringEffect_4BPmin_GENES$Name %in% DMS_PaternalEffect_4BPmin_GENES$Name & 
                                              !DMS_OffspringEffect_4BPmin_GENES$Name %in% setDMSInversionSlope_GENES$Name,]
annot_caseVennG1G2NOinter = DMS_OffspringEffect_4BPmin_GENES[DMS_OffspringEffect_4BPmin_GENES$Name %in% DMS_PaternalEffect_4BPmin_GENES$Name & 
                                                   !DMS_OffspringEffect_4BPmin_GENES$Name %in% setDMSInversionSlope_GENES$Name,]
annot_caseVennG1G2inter = DMS_OffspringEffect_4BPmin_GENES[DMS_OffspringEffect_4BPmin_GENES$Name %in% DMS_PaternalEffect_4BPmin_GENES$Name & 
                                                 DMS_OffspringEffect_4BPmin_GENES$Name %in% setDMSInversionSlope_GENES$Name,] 
annot_caseVennG2interNOG1 = DMS_OffspringEffect_4BPmin_GENES[!DMS_OffspringEffect_4BPmin_GENES$Name %in% DMS_PaternalEffect_4BPmin_GENES$Name &
                                                   DMS_OffspringEffect_4BPmin_GENES$Name %in% setDMSInversionSlope_GENES$Name,]
```

### Save annotated Manhattan plots

```{r manhattanCasesVenn,fig.height = 5, fig.width = 16}
namesCases = c("G1", "G2", "G1+G2", "G1:G2", "G2noG1")
casesVec=c("annot_caseVennG1only", "annot_caseVennG2only", "annot_caseVennG1G2NOinter", "annot_caseVennG1G2inter", "annot_caseVennG2interNOG1")

plotAndsave <- function(i){
  P=plotManhattanGenesDMS4BP(annotFile = get(casesVec[i]),
                           GYgynogff = GYgynogff, isBPinfo = FALSE, 
                           myxlab = paste0("Genes linked with the ", namesCases[[i]], " effect"))
  print(P)
  pdf(paste0("../../dataOut/Manhattan", namesCases[[i]], ".pdf"), width = 16, height = 5)
  P
}

annot_caseVennG1G2inter[annot_caseVennG1G2inter$chrom %in% "Gy_chrI",]

plotAndsave(1);  dev.off()
plotAndsave(2);  dev.off()
plotAndsave(3);  dev.off()
plotAndsave(4);  dev.off()
plotAndsave(5);  dev.off()
rm(casesVec,namesCases, plotAndsave)
```

### Gene Ontology analysis:

-   Gene universe: all genes which are present in the dataset
-   Gene sub-universe: all genes in DMS in the GOterm universe we only
want genes with (1) DMS between G2 exposed to parasites, by brother
pair, present in AT LEAST 4 BP/8 for which (2) CpGs were covered,
and (3) which have a GOterm attributed

```{r geneOnto_createUniverse}
# create gene universe
gene_universe <- data.frame(
  subsetByOverlaps(GRanges(annotGff3), GRanges(uniteCov14_G2_woSexAndUnknowChrOVERLAP))) %>% # subselect covered CpGs
  filter(lengths(Ontology_term)!=0) %>% # rm non existing GO terms
  filter(type %in% "gene")  %>% # keep all the 7416 genes with GO terms
  dplyr::select(c("Name", "Ontology_term")) %>%
  mutate(go_linkage_type = "IEA") %>% #NB: IEA but not necessarily true, it's from Interproscan after Maker. Sticklebacks (biomart) have 82701 IEA and 63 ISS.
  relocate("Ontology_term","go_linkage_type","Name") %>%
  unnest(Ontology_term) %>% # one GO per line (was a list before in this column)
  data.frame()

# Create gene set collection
goFrame <- GOFrame(gene_universe, organism="Gasterosteus aculeatus")
goAllFrame <- GOAllFrame(goFrame)
gsc_universe <- GeneSetCollection(goAllFrame, setType = GOCollection())
```

**IMPORTANT NOTE from Mel: why conditional hypergeometric test?** The GO ontology is set up as a directed acyclic graph, where a parent term is comprised of all its child terms. If you do a standard hypergeometric, you might e.g., find 'positive regulation of kinase activity' to be significant. If you then test 'positive regulation of catalytic activity', which is a parent term, then it might be significant as well, but only because of the terms coming from positive regulation of kinase activity.

The conditional hypergeometric takes this into account, and only uses those terms that were not already significant when testing a higher order (parent) term.

```{r geneOnto_runGO}
GO_G1only = makedfGO(annot_caseVennG1only, gene_universe, effect = "G1only")
GO_G2only = makedfGO(annot_caseVennG2only, gene_universe, effect = "G2only")
GO_G1G2NOinter = makedfGO(annot_caseVennG1G2NOinter, gene_universe, effect = "G1G2NOinter")
GO_G1G2inter = makedfGO(annot_caseVennG1G2inter, gene_universe, effect = "G1G2inter")
GO_G2interNOG1 = makedfGO(annot_caseVennG2interNOG1, gene_universe, effect = "G2interNOG1")

dfGO = rbind(GO_G1only, GO_G2only, GO_G1G2NOinter, GO_G1G2inter, GO_G2interNOG1)
```

### GO plot

```{r geneOnto_plotGO, fig.height = 25, fig.width = 10}
dfGO %>% ggplot(aes(x=Effect, y = factor(GO.name))) +
  geom_point(aes(color = p.value.adjusted, size = genePercent)) +
  scale_color_gradient(name="adjusted\np-value", low = "red", high = "blue") +
  scale_size_continuous(name = "% of genes")+
  theme_bw() + ylab("") + xlab("Treatments comparison") +
  theme(legend.box.background = element_rect(fill = "#ebebeb", color = "#ebebeb"),
        legend.background = element_rect(fill = "#ebebeb", color = "#ebebeb"),
        legend.key = element_rect(fill = "#ebebeb", color = "#ebebeb"), legend.position="left") + # grey box for legend
  facet_grid(fct_inorder(GO.category)~., scales="free",space = "free")
```

## Correlation between methylation and phenotype (nbr worms, BCI)

Approach:

1. Extract methylation values

-   raw beta values at DMS shared by \>4 (or more) BP or
-   derive difference between individual and mean of the group

2.  PCA

3.  extract axes 1 & 2

4.  Correlation with parasite load/BCI

```{r BCI}
# PARENTAL effect: DMS found in either CC-TC or CT-TT comparisons
# DMS_PaternalEffect_4BPmin
# 
# OFFSPRING effect: DMS found in either CC-CT or TC-TT comparisons
# DMS_OffspringEffect_4BPmin
#
# INTERACTION effect: DMS found in CC-CT which show a differential methylation (not necessarily significant) 
# in the opposite direction in TC-TT, or inversely (reaction norm are inversed)
# setDMSInversionSlope

# The five cases of Venn diagram:
# caseVennG1only
# caseVennG2only              
# caseVennG1G2NOinter
# caseVennG1G2inter
# caseVennG2interNOG1

### I. PCA based on all methylation values at DMS positions detected for INTERACTION effect, with imputation of missing values

# 1. get raw values
pos2keep = which(paste(uniteCov14_G2_woSexAndUnknowChrOVERLAP$chr, uniteCov14_G2_woSexAndUnknowChrOVERLAP$start, sep = " ") %in% 
                   caseVennG1G2inter)
uniteAtDMS = methylKit::select(uniteCov14_G2_woSexAndUnknowChrOVERLAP, pos2keep)
percAtDMS = percMethylation(uniteAtDMS)

nrow(percAtDMS) # 190 DMS linked with INTERACTION effect G1:G2

# We use missMDA and FactoMineR for imputation of missing data and
# performing of PCA: (see:
# 
#                       -   <http://juliejosse.com/wp-content/uploads/2018/05/DataAnalysisMissingR.html>
#                       -   <https://www.youtube.com/watch?v=OOM8_FH6_8o>)

PCA_percAtDMS_imputed <- myPCA(x = t(percAtDMS), incomplete = TRUE)

### PCA plot of individuals with associated colors for paternal treatments
fviz_pca_ind(PCA_percAtDMS_imputed$res.PCA,  label="none", habillage=as.factor(PCA_percAtDMS_imputed$metadata$PAT)) +
  scale_shape_manual(values=c(19,19))

### PCA plot of individuals with associated colors for brother pair
fviz_pca_ind(PCA_percAtDMS_imputed$res.PCA,  label="none", habillage=as.factor(PCA_percAtDMS_imputed$metadata$brotherPairID))

# The function dimdesc() can be used to identify the most correlated variables with a given principal component.
mydimdesc <- dimdesc(PCA_percAtDMS_imputed$res.PCA, axes = c(1,2), proba = 0.05)

nrow(mydimdesc$Dim.1$quanti)# 97 CpG sites most correlated (p < 0.05) with the first principal component 
nrow(mydimdesc$Dim.2$quanti)# 76 with the second principal component

formula(PCA_percAtDMS_imputed$modSel)
# The SECOND PCA axis is significant in BCI
# BCI ~ PCA2 + No.Worms + PAT + (1 | brotherPairID) + (1 | Sex) + 
#   PCA2:PAT + No.Worms:PAT

### How much of the BCI variance is explained by each variables?
mod_noworms = lmer(BCI ~ PCA2 + PAT + PCA2:PAT + (1 | brotherPairID) + (1 | Sex), data = PCA_percAtDMS_imputed$metadata)
mod_noPAT = lmer(BCI ~ PCA2 + No.Worms + (1 | brotherPairID) + (1 | Sex), data = PCA_percAtDMS_imputed$metadata)
mod_noPCA2 = lmer(BCI ~ + No.Worms + PAT + No.Worms:PAT +(1 | brotherPairID) + (1 | Sex), data = PCA_percAtDMS_imputed$metadata)

# R2c conditional R2 value associated with fixed effects plus the random effects.
A = (MuMIn::r.squaredGLMM(PCA_percAtDMS_imputed$modSel)[2] -
       MuMIn::r.squaredGLMM(mod_noworms)[2])*100
B = (MuMIn::r.squaredGLMM(PCA_percAtDMS_imputed$modSel)[2] -
       MuMIn::r.squaredGLMM(mod_noPAT)[2])*100
C = (MuMIn::r.squaredGLMM(PCA_percAtDMS_imputed$modSel)[2] -
       MuMIn::r.squaredGLMM(mod_noPCA2)[2])*100

round(A, 2) #% of the variance in associated with the parasite load (number of worms)
round(B, 2) #% of the variance in associated with the paternal treatment
round(C, 2) #% of the variance in associated with the first PCA axis

### Plot of the model
phenoMethPlot1 <- plot(ggpredict(PCA_percAtDMS_imputed$modSel, terms = c("No.Worms", "PCA2", "PAT")), add.data = TRUE, alpha = .08) + theme_bw() +
  scale_color_gradient(low = "blue", high = "red")+
  scale_fill_gradient(low = "blue", high = "red") +
  ylab("Body Condition Index") + xlab("Number of worms")+
  ggtitle("Predicted values of Body Condition Index in offspring")

phenoMethPlot1

# save
pdf(file = "../../dataOut/phenoMethPlot_G1:G2effect.pdf", width = 10, height = 5)
phenoMethPlot1
dev.off()

##################################
### II. PCA based on all methylation values at DMS positions detected for PATERNAL effect, with imputation of missing values

# 1. get raw values
pos2keep = which(paste(uniteCov14_G2_woSexAndUnknowChrOVERLAP$chr, uniteCov14_G2_woSexAndUnknowChrOVERLAP$start, sep = " ") %in% 
                   caseVennG1only)
uniteAtDMS = methylKit::select(uniteCov14_G2_woSexAndUnknowChrOVERLAP, pos2keep)
percAtDMS = percMethylation(uniteAtDMS)

nrow(percAtDMS) # DMS linked with parental effect only

# We use missMDA and FactoMineR for imputation of missing data and
# performing of PCA: (see:
# 
#                       -   <http://juliejosse.com/wp-content/uploads/2018/05/DataAnalysisMissingR.html>
#                       -   <https://www.youtube.com/watch?v=OOM8_FH6_8o>)

PCA_percAtDMS_imputed <- myPCA(x = t(percAtDMS), incomplete = TRUE)

### PCA plot of individuals with associated colors for paternal treatments
fviz_pca_ind(PCA_percAtDMS_imputed$res.PCA,  label="none", habillage=as.factor(PCA_percAtDMS_imputed$metadata$PAT)) +
  scale_shape_manual(values=c(19,19))

### PCA plot of individuals with associated colors for brother pair
fviz_pca_ind(PCA_percAtDMS_imputed$res.PCA,  label="none", habillage=as.factor(PCA_percAtDMS_imputed$metadata$brotherPairID))

# The function dimdesc() can be used to identify the most correlated variables with a given principal component.
mydimdesc <- dimdesc(PCA_percAtDMS_imputed$res.PCA, axes = c(1,2), proba = 0.05)

nrow(mydimdesc$Dim.1$quanti)# CpG sites most correlated (p < 0.05) with the first principal component 
nrow(mydimdesc$Dim.2$quanti)# with the second principal component.

formula(PCA_percAtDMS_imputed$modSel)
# The second PCA axis is significant in BCI
# BCI ~ PCA2 + No.Worms + PAT + (1 | brotherPairID) + (1 | Sex) + 
#   PCA2:PAT + No.Worms:PAT

### How much of the BCI variance is explained by each variables?
mod_noworms = lmer(BCI ~ PCA2 + PAT + PCA2:PAT + (1 | brotherPairID) + (1 | Sex), data = PCA_percAtDMS_imputed$metadata)
mod_noPAT = lmer(BCI ~ PCA2 + No.Worms + (1 | brotherPairID) + (1 | Sex), data = PCA_percAtDMS_imputed$metadata)
mod_noPCA2 = lmer(BCI ~ No.Worms + PAT + No.Worms:PAT + (1 | brotherPairID) + (1 | Sex), data = PCA_percAtDMS_imputed$metadata)

# R2c conditional R2 value associated with fixed effects plus the random effects.
A = (MuMIn::r.squaredGLMM(PCA_percAtDMS_imputed$modSel)[2] -
       MuMIn::r.squaredGLMM(mod_noworms)[2])*100
B = (MuMIn::r.squaredGLMM(PCA_percAtDMS_imputed$modSel)[2] -
       MuMIn::r.squaredGLMM(mod_noPAT)[2])*100
C = (MuMIn::r.squaredGLMM(PCA_percAtDMS_imputed$modSel)[2] -
       MuMIn::r.squaredGLMM(mod_noPCA2)[2])*100

round(A, 2) #% of the variance in associated with the parasite load (number of worms)
round(B, 2) #% of the variance in associated with the paternal treatment
round(C, 2) #% of the variance in associated with the second PCA axis

### Plot of the model
phenoMethPlot <- plot(ggpredict(PCA_percAtDMS_imputed$modSel, terms = c("No.Worms", "PCA2", "PAT")), add.data = TRUE, alpha = .08) + theme_bw() +
  scale_color_gradient(low = "blue", high = "red")+
  scale_fill_gradient(low = "blue", high = "red") +
  ylab("Body Condition Index") + xlab("Number of worms") +
  ggtitle("Predicted values of Body Condition Index in offspring")

phenoMethPlot

# save
pdf(file = "../../dataOut/phenoMethPlot_G1effectonly.pdf", width = 10, height = 5)
phenoMethPlot
dev.off()

##################################
### III. PCA based on all methylation values at DMS positions detected for OFFSPRING effect, with imputation of missing values

# 1. get raw values
pos2keep = which(paste(uniteCov14_G2_woSexAndUnknowChrOVERLAP$chr, uniteCov14_G2_woSexAndUnknowChrOVERLAP$start, sep = " ") %in% 
                   caseVennG2only)
uniteAtDMS = methylKit::select(uniteCov14_G2_woSexAndUnknowChrOVERLAP, pos2keep)
percAtDMS = percMethylation(uniteAtDMS)

nrow(percAtDMS) # DMS linked with offspring effect only

# We use missMDA and FactoMineR for imputation of missing data and
# performing of PCA: (see:
# 
#                       -   <http://juliejosse.com/wp-content/uploads/2018/05/DataAnalysisMissingR.html>
#                       -   <https://www.youtube.com/watch?v=OOM8_FH6_8o>)

PCA_percAtDMS_imputed <- myPCA(x = t(percAtDMS), incomplete = TRUE)

### PCA plot of individuals with associated colors for paternal treatments
fviz_pca_ind(PCA_percAtDMS_imputed$res.PCA,  label="none", habillage=as.factor(PCA_percAtDMS_imputed$metadata$PAT)) +
  scale_shape_manual(values=c(19,19))

### PCA plot of individuals with associated colors for brother pair
fviz_pca_ind(PCA_percAtDMS_imputed$res.PCA,  label="none", habillage=as.factor(PCA_percAtDMS_imputed$metadata$brotherPairID))

# The function dimdesc() can be used to identify the most correlated variables with a given principal component.
mydimdesc <- dimdesc(PCA_percAtDMS_imputed$res.PCA, axes = c(1,2), proba = 0.05)

nrow(mydimdesc$Dim.1$quanti)# CpG sites most correlated (p < 0.05) with the first principal component 
nrow(mydimdesc$Dim.2$quanti)# with the second principal component.

formula(PCA_percAtDMS_imputed$modSel)
# The first PCA axis is significant in BCI
# BCI ~ PCA1 + No.Worms + PAT + (1 | brotherPairID) + (1 | Sex) + 
#   PCA1:PAT + No.Worms:PAT

### How much of the BCI variance is explained by each variables?
mod_noworms = lmer(BCI ~ PCA1 + PAT + PCA1:PAT + (1 | brotherPairID) + (1 | Sex), data = PCA_percAtDMS_imputed$metadata)
mod_noPAT = lmer(BCI ~ PCA1 + No.Worms + (1 | brotherPairID) + (1 | Sex), data = PCA_percAtDMS_imputed$metadata)
mod_noPCA1 = lmer(BCI ~ No.Worms + PAT + No.Worms:PAT + (1 | brotherPairID) + (1 | Sex), data = PCA_percAtDMS_imputed$metadata)

# R2c conditional R2 value associated with fixed effects plus the random effects.
A = (MuMIn::r.squaredGLMM(PCA_percAtDMS_imputed$modSel)[2] -
       MuMIn::r.squaredGLMM(mod_noworms)[2])*100
B = (MuMIn::r.squaredGLMM(PCA_percAtDMS_imputed$modSel)[2] -
       MuMIn::r.squaredGLMM(mod_noPAT)[2])*100
C = (MuMIn::r.squaredGLMM(PCA_percAtDMS_imputed$modSel)[2] -
       MuMIn::r.squaredGLMM(mod_noPCA2)[2])*100

round(A, 2) #% of the variance in associated with the parasite load (number of worms)
round(B, 2) #% of the variance in associated with the paternal treatment
round(C, 2) #% of the variance in associated with the second PCA axis

### Plot of the model
phenoMethPlotG2 <- plot(ggpredict(PCA_percAtDMS_imputed$modSel, terms = c("No.Worms", "PCA1", "PAT")), add.data = TRUE, alpha = .08) + theme_bw() +
  scale_color_gradient(low = "blue", high = "red")+
  scale_fill_gradient(low = "blue", high = "red") +
  ylab("Body Condition Index") + xlab("Number of worms") +
  ggtitle("Predicted values of Body Condition Index in offspring")

phenoMethPlotG2

# save
pdf(file = "../../dataOut/phenoMethPlot_G2effectonly.pdf", width = 10, height = 5)
phenoMethPlot
dev.off()
```

# Differential methylations "Top down": DM in parents -\> how do they look in offspring?

### TREATMENT EFFECT: Differential methylation between control and infected, within each parental treatment

```{r}
## Features Annotation (use package genomation v1.24.0)
## NB Promoters are defined by options at genomation::readTranscriptFeatures function.
## The default option is to take -1000,+1000bp around the TSS and you can change that.
## -> following Heckwolf 2020 and Sagonas 2020, we consider 1500bp upstream and 500 bp downstream

## Parents comparison:
diffAnn_PAR = annotateWithGeneParts(as(DMSlist$DMS_15pc_G1_C_T,"GRanges"),annotBed12)
## Offspring from control parents comparison:
diffAnn_G2_controlG1 = annotateWithGeneParts(as(DMSlist$DMS_15pc_CC_CT,"GRanges"),annotBed12)
## Offspring from infected parents comparison:
diffAnn_G2_infectedG1 = annotateWithGeneParts(as(DMSlist$DMS_15pc_TC_TT,"GRanges"),annotBed12)
```

```{r, fig.height = 6, fig.width = 10}
par(mfrow=c(1,3))
par(mar = c(.1,0.1,5,0.1)) # Set the margin on all sides to 2
## Parents comparison:
diffAnn_PAR = annotateWithGeneParts(as(DMSlist$DMS_15pc_G1_C_T,"GRanges"),annotBed12)
diffAnn_PAR
genomation::plotTargetAnnotation(diffAnn_PAR,precedence=TRUE, main="DMS G1", cex.legend = 1, border="white")

## Offspring from control parents comparison:
diffAnn_G2_controlG1 = annotateWithGeneParts(as(DMSlist$DMS_15pc_CC_CT,"GRanges"),annotBed12)
diffAnn_G2_controlG1
genomation::plotTargetAnnotation(diffAnn_G2_controlG1,precedence=TRUE, main="DMS G2-G1c", cex.legend = 1, border="white")
## Offspring from infected parents comparison:
diffAnn_G2_infectedG1 = annotateWithGeneParts(as(DMSlist$DMS_15pc_TC_TT,"GRanges"),annotBed12)
diffAnn_G2_infectedG1
genomation::plotTargetAnnotation(diffAnn_G2_infectedG1,precedence=TRUE, main="DMS G2-G1i", cex.legend = 1, border="white")
par(mfrow=c(1,1))
```

```{r}
## Run the function to get DMS info
DMS_info_G1 <- myDMSinfo(DMSlist$DMS_15pc_G1_C_T, uniteCov6_G1_woSexAndUnknowChrOVERLAP)
DMS_info_G2_G1c_final <- myDMSinfo(DMSlist$DMS_15pc_CC_CT, uniteCov14_G2_woSexAndUnknowChrOVERLAP)
DMS_info_G2_G1i_final <- myDMSinfo(DMSlist$DMS_15pc_TC_TT,uniteCov14_G2_woSexAndUnknowChrOVERLAP)
```

NB Kostas' results: "We found a total of 1,973 CpG sites out of
1,172,887 CpGs (0.17%) across the genome that showed at least 15%
differential fractional methylation (differentially methylated site
[DMS]; q \< 0.01) between infected and uninfected fish"

Here: we obtain out a total of
`r nrow(uniteCov14_G2_woSexAndUnknowChrOVERLAP)` CpG sites:
`r length(DMS_info_G1$DMS)` (`r round(DMS_info_G1$percentDMS, 2)`%)
showed at least 15% differential fractional methylation (differentially
methylated site [DMS]; q \< 0.01) between infected and uninfected fish
in the parental group; `r length(DMS_info_G2_G1c_final$DMS)`
(`r round(DMS_info_G2_G1c_final$percentDMS, 2)`%) in the offspring from
**control parents** comparison; `r length(DMS_info_G2_G1i_final$DMS)`
(`r round(DMS_info_G2_G1i_final$percentDMS, 2)`%) in the offspring from
**infected parents** comparison.

#### Intersections of DMS: Venn diagrams

```{r}
## Chi2 test: are the number of DMS from G2-G1C and G2-G1I overlapping with DMSpar statistically different?

A=length(intersect(DMS_info_G1$DMS,DMS_info_G2_G1c_final$DMS))
B=length(DMS_info_G2_G1c_final$DMS)
C=length(intersect(DMS_info_G1$DMS,DMS_info_G2_G1i_final$DMS))
D=length(DMS_info_G2_G1i_final$DMS)

Observed=matrix(c(A, B-A, C, D-C),nrow=2)
Observed

chisq.test(Observed)
## not statistically different
```

```{r, fig.width = 12, fig.height = 15}
## output Venn diagrams
allVenn <- ggVennDiagram(list("DMS G1" = DMS_info_G1$DMS, "DMS G2-c" = DMS_info_G2_G1c_final$DMS, "DMS G2-i" = DMS_info_G2_G1i_final$DMS), label_alpha = 0) +
  scale_fill_gradient(low="white",high = "red")
hypoVenn <- ggVennDiagram(list("DMS G1\nhypo" = DMS_info_G1$DMS[DMS_info_G1$direction %in% "hypo"],
                               "DMS G2-c\nhypo" = DMS_info_G2_G1c_final$DMS[DMS_info_G2_G1c_final$direction %in% "hypo"],
                               "DMS G2-i\nhypo" = DMS_info_G2_G1i_final$DMS[DMS_info_G2_G1i_final$direction %in% "hypo"]), label_alpha = 0) +
  scale_fill_gradient(low="white",high = "red")
hyperVenn <- ggVennDiagram(list("DMS G1\nhyper" = DMS_info_G1$DMS[DMS_info_G1$direction %in% "hyper"],
                                "DMS G2-c\nhyper" = DMS_info_G2_G1c_final$DMS[DMS_info_G2_G1c_final$direction %in% "hyper"],
                                "DMS G2-i\nhyper" = DMS_info_G2_G1i_final$DMS[DMS_info_G2_G1i_final$direction %in% "hyper"]), label_alpha = 0) +
  scale_fill_gradient(low="white",high = "red")

ggarrange(allVenn,
          ggarrange(hypoVenn, hyperVenn, ncol = 2, legend = "none"),
          nrow = 2, widths = c(.5,1))
```

##### Venn with annotated features

```{r}
runHyperHypoAnnot <- function(){
  par(mfrow=c(2,3))
  par(mar = c(.1,0.1,5,0.1)) # Set the margin on all sides to 2
  ####### HYPO
  ## Parents comparison:
  A = annotateWithGeneParts(
    as(DMSlist$DMS_15pc_G1_C_T[DMS_info_G1$direction %in% "hypo",],"GRanges"),annotBed12)
  genomation::plotTargetAnnotation(A,precedence=TRUE, main="DMS G1\nhypo",
                                   cex.legend = .4, border="white")
  ## Offspring from control parents comparison:
  B = annotateWithGeneParts(
    as(DMSlist$DMS_15pc_CC_CT[DMS_info_G2_G1c_final$direction %in% "hypo",],"GRanges"),annotBed12)
  genomation::plotTargetAnnotation(B,precedence=TRUE, main="DMS G2-G1c\nhypo",
                                   cex.legend = .4, border="white")
  ## Offspring from infected parents comparison:
  C = annotateWithGeneParts(
    as(DMSlist$DMS_15pc_TC_TT[DMS_info_G2_G1i_final$direction %in% "hypo",],"GRanges"),annotBed12)
  genomation::plotTargetAnnotation(C,precedence=TRUE, main="DMS G2-G1i\nhypo",
                                   cex.legend = .4, border="white")
  ####### HYPER
  ## Parents comparison:
  D = annotateWithGeneParts(
    as(DMSlist$DMS_15pc_G1_C_T[DMS_info_G1$direction %in% "hyper",],"GRanges"),annotBed12)
  genomation::plotTargetAnnotation(D,precedence=TRUE, main="DMS G1\nhyper",
                                   cex.legend = .4, border="white")
  ## Offspring from control parents comparison:
  E = annotateWithGeneParts(
    as(DMSlist$DMS_15pc_CC_CT[DMS_info_G2_G1c_final$direction %in% "hyper",],"GRanges"),annotBed12)
  genomation::plotTargetAnnotation(E,precedence=TRUE, main="DMS G2-G1c\nhyper",
                                   cex.legend = .4, border="white")
  ## Offspring from infected parents comparison:
  f = annotateWithGeneParts(
    as(DMSlist$DMS_15pc_TC_TT[DMS_info_G2_G1i_final$direction %in% "hyper",],"GRanges"),annotBed12)
  genomation::plotTargetAnnotation(f,precedence=TRUE, main="DMS G2-G1i\nhyper",
                                   cex.legend = .4, border="white")
  par(mfrow=c(1,1))
  return(list(G1hypo=A, G2G1chypo=B, G2G1ihypo=C, G1hyper=D, G2G1chyper=E, G2G1ihyper=f))
}

myannot=runHyperHypoAnnot()
```

```{r}
############################################################
## Venn diagram of overlapping features by their annotation:
table(rowSums(as.data.frame(myannot$G1hypo@members))) # NB: some positions are labelled with several features!
## as in MBE 2021: "giving precedence to the following order promoters, exons,
## introns, and intergenic regions when features overlapped"

myAnnotateDMS <- function(DMS, annot){
  ## sanity check
  if (nrow(DMS) != nrow(annot)){"STOP error in arguments"}
  DMS$pos <- paste(DMS$chr, DMS$start, DMS$end)
  ## NB as in MBE 2021: "giving precedence to the following order promoters, exons,
  ## introns, and intergenic regions when features overlapped"
  DMS$feature <- NA
  ## 1. promoters
  DMS$feature[which(annot$prom == 1)] = "promoter"
  ## 2. exons
  DMS$feature[which(annot$exon == 1 & annot$prom ==0)] = "exon"
  ## 3. intron
  DMS$feature[which(annot$intro == 1 & annot$exon == 0 & annot$prom ==0)] = "intron"
  ## 4. intergenic regions
  DMS$feature[which(annot$intro == 0 & annot$exon == 0 & annot$prom ==0)] = "intergenic"
  return(DMS)
}

DMSlist$DMS_15pc_G1_C_T = myAnnotateDMS(DMSlist$DMS_15pc_G1_C_T, as.data.frame(diffAnn_PAR@members))
DMSlist$DMS_15pc_G1_C_T_HYPO = myAnnotateDMS(DMSlist$DMS_15pc_G1_C_T[DMS_info_G1$direction %in% "hypo",],
                                             as.data.frame(myannot$G1hypo@members))
DMSlist$DMS_15pc_G1_C_T_HYPER = myAnnotateDMS(DMSlist$DMS_15pc_G1_C_T[DMS_info_G1$direction %in% "hyper",],
                                              as.data.frame(myannot$G1hyper@members))

DMSlist$DMS_15pc_CC_CT = myAnnotateDMS(DMSlist$DMS_15pc_CC_CT, as.data.frame(diffAnn_G2_controlG1@members))
DMSlist$DMS_15pc_CC_CT_HYPO = myAnnotateDMS(DMSlist$DMS_15pc_CC_CT[DMS_info_G2_G1c_final$direction %in% "hypo",],
                                            as.data.frame(myannot$G2G1chypo@members))
DMSlist$DMS_15pc_CC_CT_HYPER = myAnnotateDMS(DMSlist$DMS_15pc_CC_CT[DMS_info_G2_G1c_final$direction %in% "hyper",],
                                             as.data.frame(myannot$G2G1chyper@members))

DMSlist$DMS_15pc_TC_TT = myAnnotateDMS(DMSlist$DMS_15pc_TC_TT, as.data.frame(diffAnn_G2_infectedG1@members))
DMSlist$DMS_15pc_TC_TT_HYPO = myAnnotateDMS(DMSlist$DMS_15pc_TC_TT[DMS_info_G2_G1i_final$direction %in% "hypo",],
                                            as.data.frame(myannot$G2G1ihypo@members))
DMSlist$DMS_15pc_TC_TT_HYPER = myAnnotateDMS(DMSlist$DMS_15pc_TC_TT[DMS_info_G2_G1i_final$direction %in% "hyper",],
                                             as.data.frame(myannot$G2G1ihyper@members))

## Make Venn diagram for each feature
getFeatureDFHYPO <- function(myfeat){
  a = DMSlist$DMS_15pc_G1_C_T_HYPO$pos[DMSlist$DMS_15pc_G1_C_T_HYPO$feature %in% myfeat]
  b = DMSlist$DMS_15pc_CC_CT_HYPO$pos[DMSlist$DMS_15pc_CC_CT_HYPO$feature %in% myfeat]
  c = DMSlist$DMS_15pc_TC_TT_HYPO$pos[DMSlist$DMS_15pc_TC_TT_HYPO$feature %in% myfeat]
  return(list(a=a,b=b,c=c))
}

getFeatureDFHYPER <- function(myfeat){
  a = DMSlist$DMS_15pc_G1_C_T_HYPER$pos[DMSlist$DMS_15pc_G1_C_T_HYPER$feature %in% myfeat]
  b = DMSlist$DMS_15pc_CC_CT_HYPER$pos[DMSlist$DMS_15pc_CC_CT_HYPER$feature %in% myfeat]
  c = DMSlist$DMS_15pc_TC_TT_HYPER$pos[DMSlist$DMS_15pc_TC_TT_HYPER$feature %in% myfeat]
  return(list(a=a,b=b,c=c))
}

getVenn <- function(feat, direction){
  if (direction == "hypo"){
    ggVennDiagram(list(A = getFeatureDFHYPO(feat)[["a"]],
                       B = getFeatureDFHYPO(feat)[["b"]],
                       C = getFeatureDFHYPO(feat)[["c"]]), label_alpha = 0,
                  category.names = c(paste0("DMS G1\nhypo\n", feat), paste0("DMS G2-c\nhypo\n", feat), paste0("DMS G2-i\nhypo\n", feat))) +
      scale_fill_gradient(low="white",high = "red")
  } else if (direction == "hyper"){
    ggVennDiagram(list(A = getFeatureDFHYPER(feat)[["a"]],
                       B = getFeatureDFHYPER(feat)[["b"]],
                       C = getFeatureDFHYPER(feat)[["c"]]), label_alpha = 0,
                  category.names = c(paste0("DMS G1\nhyper\n", feat), paste0("DMS G2-c\nhyper\n", feat), paste0("DMS G2-i\nhyper\n", feat))) +
      scale_fill_gradient(low="white",high = "red")
  }
}
```

```{r, fig.width = 12, fig.height = 15}
ggarrange(getVenn("promoter", "hypo"), getVenn("exon", "hypo"),
          getVenn("intron", "hypo"), getVenn("intergenic", "hypo"),
          nrow = 2, ncol = 2)
```

```{r, fig.width = 12, fig.height = 15}
ggarrange(getVenn("promoter", "hyper"), getVenn("exon", "hyper"),
          getVenn("intron", "hyper"), getVenn("intergenic", "hyper"),
          nrow = 2, ncol = 2)
```

#### Manhattan plot of DMS

```{r, fig.width= 10, fig.height=3}
## Parents trt-ctrl
# load annotation
annot_PAR <- as.data.frame(diffAnn_PAR@members)
makeManhattanPlots(DMSfile = DMSlist$DMS_15pc_G1_C_T, annotFile = annot_PAR, GYgynogff = GYgynogff,
                   mycols = c("red", "grey", "black", "green"), mytitle = "Manhattan plot of G1 DMS")

## G2-G1c trt-ctrl
# load annotation
annot_G2_G1c <- as.data.frame(diffAnn_G2_controlG1@members)
makeManhattanPlots(DMSfile = DMSlist$DMS_15pc_CC_CT, annotFile = annot_G2_G1c, GYgynogff = GYgynogff,
                   mycols = c("red", "grey", "black", "green"), mytitle = "Manhattan plot of G2-G1c DMS")

## G2-G1i trt-ctrl
# load annotation
annot_G2_G1i <- as.data.frame(diffAnn_G2_infectedG1@members)
makeManhattanPlots(DMSfile = DMSlist$DMS_15pc_TC_TT, annotFile = annot_G2_G1i, GYgynogff = GYgynogff,
                   mycols = c("red", "grey", "black", "green"), mytitle = "Manhattan plot of G2-G1i DMS")

## Outliers in Manhattan plot: 15% diff + 2SD
outliers_G1_final <- which(abs(DMSlist$DMS_15pc_G1_C_T$meth.diff) > 15 + 2*sd(abs(DMSlist$DMS_15pc_G1_C_T$meth.diff)))
outliers_annot_G1 <- as.data.frame(diffAnn_PAR@members)[outliers_G1_final,]
makeManhattanPlots(DMSfile = DMSlist$DMS_15pc_G1_C_T[outliers_G1_final, ],
                   annotFile = outliers_annot_G1, GYgynogff = GYgynogff,
                   mycols = c("red", "grey", "black", "green"), mytitle = "Manhattan plot of G1 DMS")

outliers_G2_G1c_final <- which(abs(DMSlist$DMS_15pc_CC_CT$meth.diff) > 15 + 2*sd(abs(DMSlist$DMS_15pc_CC_CT$meth.diff)))
outliers_annot_G2_G1c <- as.data.frame(diffAnn_G2_controlG1@members)[outliers_G2_G1c_final,]
makeManhattanPlots(DMSfile = DMSlist$DMS_15pc_CC_CT[outliers_G2_G1c_final, ],
                   annotFile = outliers_annot_G2_G1c, GYgynogff = GYgynogff,
                   mycols = c("red", "grey", "black", "green"), mytitle = "Manhattan plot of G2-G1c DMS")

outliers_G2_G1i_final <- which(abs(DMSlist$DMS_15pc_TC_TT$meth.diff) > 15 + 2*sd(abs(DMSlist$DMS_15pc_TC_TT$meth.diff)))
outliers_annot_G2_G1i <- as.data.frame(diffAnn_G2_infectedG1@members)[outliers_G2_G1i_final,]
makeManhattanPlots(DMSfile = DMSlist$DMS_15pc_TC_TT[outliers_G2_G1i_final, ],
                   annotFile = outliers_annot_G2_G1i, GYgynogff = GYgynogff,
                   mycols = c("red", "grey", "black", "green"), mytitle = "Manhattan plot of G2-G1i DMS")
```

## I. Focus on CpG positions at parental (Ctrl-Inf) DMS

Question: how are the beta values in the different groups at the
parental DMS?

```{r}
##############
## Prepare dataset
PM_G1 <- getPMdataset(uniteCov = uniteCov6_G1_woSexAndUnknowChrOVERLAP, MD = fullMetadata_PAR, gener="parents")
PM_G2 <- getPMdataset(uniteCov = uniteCov14_G2_woSexAndUnknowChrOVERLAP, MD = fullMetadata_OFFS, gener="offspring")
```

### What is the relative contribution of methylation to brother pair & paternal treatment?

Test of VCA: variance component analysis
<https://cran.r-project.org/web/packages/VCA/vignettes/VCA_package_vignette.html>

#### Hypo methylation

```{r, fig.width=10}
PM_G2_mean_hypo <- PM_G2[PM_G2$hypohyper %in% "hypo", ] %>%
  group_by(brotherPairID, G1_trt, G2_trt, ID) %>%
  dplyr::summarize(MeanBetaValue = mean(BetaValue, na.rm=TRUE)) %>% data.frame()

varPlot(form = MeanBetaValue~(G1_trt* G2_trt*brotherPairID), Data = PM_G2_mean_hypo,
        MeanLine=list(var=c("G1_trt", "G2_trt"),
                      col=c("white", "blue"), lwd=c(2,2)),
        BG=list(var="G2_trt", col=paste0("gray", c(80, 90))),
        YLabel=list(cex = .8, text="Mean beta value at parDMS \n hypomethylated upon infection"))
```

```{r}
myfitVCA_hypo <- fitVCA(form = MeanBetaValue~(G1_trt* G2_trt*brotherPairID), Data = PM_G2_mean_hypo)

### Real values
trtEffect <- sum(myfitVCA_hypo$aov.tab[2:4, 5])
genEffect <- sum(myfitVCA_hypo$aov.tab[5:8, 5])
error <- sum(myfitVCA_hypo$aov.tab[9, 5])
realValHypoVCA <- data.frame(trtEffect=trtEffect, genEffect=genEffect,error=error)

### Randomisation
myrandomVCA <- function(df=PM_G2_mean_hypo){
  randomDF = df
  randomDF$G1_trt = sample(PM_G2_mean_hypo$G1_trt, replace = F)
  randomDF$G2_trt = sample(PM_G2_mean_hypo$G2_trt, replace = F)
  randomDF$brotherPairID = sample(PM_G2_mean_hypo$brotherPairID, replace = F)
  myfitVCA <- fitVCA(form = MeanBetaValue~(G1_trt* G2_trt*brotherPairID), Data = randomDF)
  trtEffect <- sum(myfitVCA$aov.tab[2:4, 5])
  genEffect <- sum(myfitVCA$aov.tab[5:8, 5])
  error <- sum(myfitVCA$aov.tab[9, 5])
  return(data.frame(trtEffect=trtEffect, genEffect=genEffect,error=error))
}

# randomHypoVCA = do.call(rbind, lapply(1:1000, function(x) {
#   df=myrandomVCA(PM_G2_mean_hypo)
#   df$rep=x
#   return(df)}))

# randomHypoVCA = melt(randomHypoVCA, id.vars = "rep")
# saveRDS(randomHypoVCA, file = "Rdata/randomHypoVCA.RDS")
randomHypoVCA <- readRDS(file = "Rdata/randomHypoVCA.RDS")
df2=reshape2::melt(realValHypoVCA)

sumDF <- randomHypoVCA %>%
  group_by(variable) %>%
  dplyr::summarize(value = mean(value)) %>% data.frame()

ggplot(randomHypoVCA, aes(x=variable, y=value))+
  geom_boxplot()+
  geom_jitter(width=.1, alpha=.2)+
  geom_point(data = df2, col = "red", size = 6)+
  geom_text(data=sumDF, aes(label=round(value)), col="white")+
  geom_text(data = df2, aes(label=round(value)), col="white")+
  theme_cleveland()+
  ggtitle("VCA with bootstrap N=1000 at hypo-parDMS", subtitle = "red: observed values")

# estimate 95% confidence intervals, request CI for
# all variance components via 'VarVC=TRUE'
VCAinference(myfitVCA_hypo, VarVC=TRUE)
```

#### Hyper methylation

```{r, fig.width=10}
PM_G2_mean_hyper <- PM_G2[PM_G2$hypohyper %in% "hyper", ] %>%
  group_by(brotherPairID, G1_trt, G2_trt, ID) %>%
  dplyr::summarize(MeanBetaValue = mean(BetaValue, na.rm=TRUE)) %>% data.frame()

varPlot(form = MeanBetaValue~(G1_trt* G2_trt*brotherPairID), Data = PM_G2_mean_hyper,
        MeanLine=list(var=c("G1_trt", "G2_trt"),
                      col=c("white", "blue"), lwd=c(2,2)),
        BG=list(var="G2_trt", col=paste0("gray", c(80, 90))),
        YLabel=list(cex = .8, text="Mean beta value at parDMS \n hypermethylated upon infection"))
```

```{r}
myfitVCA_hyper <- fitVCA(form = MeanBetaValue~(G1_trt* G2_trt*brotherPairID), Data = PM_G2_mean_hyper)

### Real values
trtEffect <- sum(myfitVCA_hyper$aov.tab[2:4, 5])
genEffect <- sum(myfitVCA_hyper$aov.tab[5:8, 5])
error <- sum(myfitVCA_hyper$aov.tab[9, 5])
realValHyperVCA <- data.frame(trtEffect=trtEffect, genEffect=genEffect,error=error)

### Randomisation
# randomHyperVCA = do.call(rbind, lapply(1:1000, function(x) {
#   df=myrandomVCA(PM_G2_mean_hyper)
#   df$rep=x
#   return(df)}))
#
# randomHyperVCA = melt(randomHyperVCA, id.vars = "rep")
# saveRDS(randomHyperVCA, file = "Rdata/randomHyperVCA.RDS")
randomHyperVCA <- readRDS(file = "Rdata/randomHyperVCA.RDS")
df2=reshape2::melt(realValHyperVCA)

sumDF <- randomHyperVCA %>%
  group_by(variable) %>%
  dplyr::summarize(value = mean(value)) %>% data.frame()

ggplot(randomHyperVCA, aes(x=variable, y=value))+
  geom_boxplot()+
  geom_jitter(width=.1, alpha=.2)+
  geom_point(data = df2, col = "red", size = 6)+
  geom_text(data=sumDF, aes(label=round(value)), col="white")+
  geom_text(data = df2, aes(label=round(value)), col="white")+
  theme_cleveland()+
  ggtitle("VCA with bootstrap N=1000 at hyper-parDMS", subtitle = "red: observed values")

# estimate 95% confidence intervals, request CI for
# all variance components via 'VarVC=TRUE'
VCAinference(myfitVCA_hypo, VarVC=TRUE)
```

## DMS values in parents

```{r}
parmod <- lmer(data = PM_G1, BetaValue ~ meth.diff.parentals : Treatment + (1|CpGSite) + (1|brotherPairID))

## check normality of residuals assumption
qqnorm(resid(parmod))
qqline(resid(parmod))

pred <- ggpredict(parmod, terms = c("meth.diff.parentals", "Treatment"))
plot(pred, add.data = T)+
  scale_color_manual(values = c("black", "red"))+
  scale_y_continuous(name = "Beta values")+
  scale_x_continuous(name = "Methylation difference between infected and control parents in percentage")+
  ggtitle("Predicted methylation ratio (Beta) values in parents\n as a function of differential methylation between exposed and control groups")+
  theme_bw()
```

### Linear model: does the beta value of offspring at DMS depends on treatment Parent x Offspring?

```{r}
modFull <- lmer(BetaValue ~ (G1_trt * G2_trt):hypohyper + (1|CpGSite) + (1|Sex) + (1|brotherPairID),data = PM_G2, REML = F) # REML =F for model comparison
mod_noG1trt <- lmer(BetaValue ~ G2_trt:hypohyper + (1|CpGSite)+ (1|Sex) + (1|brotherPairID), data = PM_G2, REML = F)
mod_noG2trt <-lmer(BetaValue ~ G1_trt:hypohyper + (1|CpGSite) + (1|Sex) + (1|brotherPairID), data = PM_G2, REML = F)
mod_noInteractions <- lmer(BetaValue ~ (G1_trt + G2_trt):hypohyper + (1|CpGSite) + (1|Sex) + (1|brotherPairID), data = PM_G2, REML = F)
mod_noHypoHyper <- lmer(BetaValue ~ (G1_trt * G2_trt) + (1|CpGSite) + (1|Sex) + (1|brotherPairID), data = PM_G2, REML = F)

## check normality of residuals assumption
qqnorm(resid(modFull))
qqline(resid(modFull))

## Likelihood ratio tests for all variables:
lrtest(modFull, mod_noG1trt) # G1 trt is VERY VERY significant (LRT:  (4) = 1163.6, p < 0.001)
lrtest(modFull, mod_noG2trt) # G2 trt is VERY VERY significant (LRT:  (4) = 30.02, p < 0.001) NB that changed when brotherpair is used instead of family!
lrtest(modFull, mod_noInteractions) # interactions are significant (LRT:  (2) = 9.21, p < 0.01)
lrtest(modFull, mod_noHypoHyper) # hypo/hyper VERY VERY significant (LRT:  (4) = 1140, p < 0.001)

## Post-hoc tests between treatments
modFull <- lmer(BetaValue ~ (G1_trt * G2_trt):hypohyper + (1|CpGSite) + (1|Sex) + (1|brotherPairID),data = PM_G2)
modFull_emmeans <- emmeans(modFull, list(pairwise ~ (G1_trt:G2_trt):hypohyper), adjust = "tukey")
modFull_emmeans

P1 <- plot(modFull_emmeans, by = "hypohyper", comparisons = TRUE) +
  # coord_flip()+
  theme_bw() +
  ggtitle("Estimated marginal means of methylation ratio (beta)\n of offspring at parental DMS")+
  theme(legend.position = "none", axis.title.x = element_blank()) +
  scale_x_continuous("Beta value (methylation ratio)", limits = c(47,69.5))

## NB: Comparison arrows: https://cran.r-project.org/web/packages/emmeans/vignettes/xplanations.html
## two estimated marginal means (EMMs) differ significantly if, and only if, their respective comparison arrows do not overlap
## These comparison arrows are decidedly not the same as confidence intervals.
## Confidence intervals for EMMs are based on the statistical properties of the individual EMMs, whereas comparison arrows
## are based on the statistical properties of differences of EMMs.

## Add the PARENTAL DMS value
## Same test on ALL, G1 and G2 fish
modFullG1 <- lmer(BetaValue ~ G1_trt:hypohyper + (1|CpGSite) + (1|brotherPairID), data = PM_G1)

modFullG1_emmeans <- emmeans(modFullG1, list(pairwise ~ G1_trt:hypohyper), adjust = "tukey")
modFullG1_emmeans

P2 <- plot(modFullG1_emmeans, by = "hypohyper", comparisons = TRUE) +
  theme_bw() +
  ggtitle("Estimated marginal means of methylation ratio (beta)\n of parents at DMS")+
  theme(legend.position = "none", axis.title.x = element_blank()) +
  scale_x_continuous("Beta value (methylation ratio)", limits = c(47,69.5))

ggarrange(P2, P1, labels = c("A", "B"), ncol = 1, nrow = 2)
```

### Are the nbr of residuals methylation AT PARENTAL DMS different in the 4 G2 trt? (for hypo vs hypermeth)?

```{r}
length(unique(PM_G1$CpGSite))# 3648 positions
PM_G1 %>% dplyr::count(ID)## NB: not all covered in all samples
length(unique(PM_G2$CpGSite[PM_G2$hypohyper %in% "hypo"]))# 1176 positions hypomethylated upon parental inf
length(unique(PM_G2$CpGSite[PM_G2$hypohyper %in% "hyper"]))# 2472 positions hypermethylated upon parental inf

myfun <- function(X){
  ## Calculate nbr of CpG hypo/hypermethylated per individual, and nbr of covered CpG:
  X <- X %>% group_by(ID, Treatment, brotherPairID, clutch.ID, Sex) %>%
    dplyr::summarise(ncov = n(),
                     hypoMeth = sum(BetaValue < 0.3),
                     hyperMeth = sum(BetaValue > 0.7)) %>% data.frame()
  # Calculate residuals of nbr of methCpG by nbr of covered CpG
  X$res_Nbr_methCpG_Nbr_coveredCpG_HYPO = residuals(lm(X$hypoMeth ~ X$ncov))
  X$res_Nbr_methCpG_Nbr_coveredCpG_HYPER = residuals(lm(X$hyperMeth ~ X$ncov))

  ## Statistical model: is it different in each offspring trt group?
  mod1 <- lmer(res_Nbr_methCpG_Nbr_coveredCpG_HYPO ~ Treatment + (1|brotherPairID/clutch.ID) + (1|Sex),
               data = X, REML = F)
  mod0 <- lmer(res_Nbr_methCpG_Nbr_coveredCpG_HYPO ~ 1 + (1|brotherPairID/clutch.ID) + (1|Sex),
               data = X, REML = F)
  print(lrtest(mod1, mod0))

  ## Post-hoc test:
  modhypo <- lmer(res_Nbr_methCpG_Nbr_coveredCpG_HYPO ~ Treatment + (1|brotherPairID/clutch.ID) + (1|Sex),
                  data = X)
  ## pairwise posthoc test
  print(emmeans(modhypo, list(pairwise ~ Treatment), adjust = "tukey"))

  mod3 <- lmer(res_Nbr_methCpG_Nbr_coveredCpG_HYPER ~ Treatment + (1|brotherPairID/clutch.ID) + (1|Sex),
               data = X, REML = F)
  mod4 <- lmer(res_Nbr_methCpG_Nbr_coveredCpG_HYPER ~ 1 + (1|brotherPairID/clutch.ID) + (1|Sex),
               data = X, REML = F)
  print(lrtest(mod3, mod4))

  ## Post-hoc test:
  modhyper <- lmer(res_Nbr_methCpG_Nbr_coveredCpG_HYPER ~ Treatment + (1|brotherPairID/clutch.ID) + (1|Sex),
                   data = X)
  ## pairwise posthoc test
  print(emmeans(modhyper, list(pairwise ~ Treatment), adjust = "tukey"))

  ## Plot
  phypo <- plot(ggpredict(modhypo, terms = c("Treatment")), add.data = TRUE)+
    scale_y_continuous("Residuals of number of hypomethylated methylated \ncytosines on number of cytosines covered") +
    ggtitle("Predicted residuals nbr of hypomethylated CpG")+
    theme_bw()

  phyper <- plot(ggpredict(modhyper, terms = c("Treatment")), add.data = TRUE)+
    scale_y_continuous("Residuals of number of hypermethylated methylated \n cytosines on number of cytosines covered") +
    ggtitle("Predicted residuals nbr of hypermethylated CpG")+
    theme_bw()
  return(list(phypo, phyper))
}

listplots <- myfun(X = PM_G2[PM_G2$hypohyper %in% "hypo",])
## NOT significant
annotate_figure(ggarrange(listplots[[1]], listplots[[2]],ncol = 2, nrow = 1),
                top = text_grob("Parental DMS hypomethylated upon infection, in offspring"))
```

```{r}
listplots <- myfun(X = PM_G2[PM_G2$hypohyper %in% "hyper",])
## Treatment SIGNIFICANT in both excess hypo/hyper methylation **

# $`pairwise differences of Treatment`
## HYPO
# 1                       estimate   SE   df t.ratio p.value
# NE_control - E_control     23.71 7.28 10.3   3.257  0.0353
# NE_control - E_exposed     26.88 7.26 10.3   3.701  0.0172

## HYPER
# 1                       estimate   SE   df t.ratio p.value
# NE_control - E_control    -24.06 7.36 10.3  -3.269  0.0348
# NE_control - E_exposed    -27.07 7.34 10.3  -3.687  0.0177

annotate_figure(ggarrange(listplots[[1]], listplots[[2]],ncol = 2, nrow = 1),
                top = text_grob("Parental DMS hypermethylated upon infection, in offspring"))
```

-\> The beta values in parentalDMS in offspring follow the parental
pattern hypo/hyper methylated upon infection
